---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 2A"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Please submit your R code and your answers. 

You can find the dataset “bpa.sas7bdat” in the R project for class this week or download it separately under "Required Activities Before Class". 

Using this dataset, do the following problems.

# Problem 1: One of your collaborators has created a SAS dataset for your study. You would like to analyze this data in R. 

## 1A. Read the .sas7bdat data into R using a function discussed in class. What are the dimensions of the bpa dataset?
```{r problem 1A}
# Insert your code for problem 1A here
```
#Interpret your answer for problem 1A here


## 1B.  Your Principal Investigator wants to look at the data in excel. Using the bpa dataset that you now have in R, export the bpa dataset to a CSV file. Now, import the new CSV back into R. What are the dimensions of the R dataset from the CSV?
```{r problem 1B}
# Insert your code for problem 1B here
```
#Interpret your answer for problem 1B here

## 1C.  C. Save the bpa dataset as an R data file (.rda). In future homework assignments, you can load the .rda file directly and will not need to use the .sas7bdat file again.
```{r problem 1C}
# Insert your code for problem 1C here
```
#Interpret your answer for problem 1C here

Problem 2. Explore the bpa dataset.

## 2A.  Compute the mean, standard deviation, minimum, 25th percentile, median, 75th percentile, and maximum of age in the bpa dataset.
```{r problem 2A}
# Insert your code for problem 2A here
```
#Interpret your answer for problem 2A here

## 2B.  Compute the mean, standard deviation, minimum, 25th percentile, median, 75th percentile, and maximum of urinary bisphenol-A (ubpa) in the bpa datset.
```{r problem 2B}
# Insert your code for problem 2B here
```
#Interpret your answer for problem 2B here

