---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Wrangling and Exploring Data with R"
author: "Kelly Bakulski, Lauren Middleton"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This is how you set options for all chunks of code

sessionInfo() # check what packages are already loaded by default
```

## Session 2, Wrangling and Exploring Data with R

# Identify base packages and install new packages

```{r install_packages, eval=FALSE}

# Install packages. Do this only once.
# Note, we already installed packages to this workspace for the class. If working on your personal computer, will need to run this code
install.packages("tidyverse")
install.packages("here")
install.packages("nhanesA")
install.packages("sjlabelled")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```

# Load packages

```{r load_packages}

# Load packages. Load relevant packages every time you start a new R session and at the top of every .Rmd file
library(tidyverse)
library(here)
library(nhanesA)
library(sjlabelled)
```

# Specify file directories

```{r directories}

here() # Orient yourself to the default file path format on your computer

#Expect to be when coding on RStudio Cloud
#"/cloud/project" 
```

# Data cleaning plan:
1) Load demographics, complete blood counts, and chemical datasets
2) Check datasets
3) Keep only the useful variables
4) Merge datasets
5) Update object types and set reference levels
6) Create categorical variables
7) Create age groups
8) Save cleaned dataset


# Import datasets from 2017-2018 NHANES
* Datasets can be found at: **https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017**
* Data are saved in multiple datasets that we will need to combine, starting with demographics
* NHANES variables have labels as well as variable names 

```{r load_demographics_data}

# Download NHANES demographics data and remove the labels
demog <- nhanes("DEMO_J")
# Check the column names (variable names)
colnames(demog)


# Check the Demographics Data website for information about these variables:
# https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Demographics&CycleBeginYear=2017
#   What does RIAGENDR mean?
```

# Explore the NHANES demographics dataset

```{r explore_data}

# Explore the data set
str(demog) # Get detailed information about each variable and the object overall
dim(demog) # What are the dimensions?
colnames(demog) # What are the column names?
rownames(demog)[1:20] # What are the row names?


head(demog) # What do the first 6 rows look like?
demog[1:10, ] # What do the first 10 rows look like?

# Look at the whole dataset
head(demog)
```


# Explore individual variables in the NHANES dataset

```{r explore_variables}

# Explore the variables, starting with RIDAGEYR
length(demog$RIDAGEYR) # How many observations are in this variable?
str(demog$RIDAGEYR) # What type of object is the variable?
demog$RIDAGEYR[1:10] # Are there any duplicates in the first few values?
```

# Check your understanding!
When you start with a new dataset, be sure to explore every variable. As an example, let's explore another variable: INDFMPIR
* Make a new R code chunk
* How long is the INDFMPIR variable?
* What type of object is the variable? (for example: character, numeric, etc.)
* What are the first few values of the variable?

# Let's load additional datasets, beyond demographics. This includes blood cell counts, metal levels, and cotinine levels

```{r load_exposure_outcome_data}

# Download Complete Blood Cell Laboratory Data, check the column names
cbc <- nhanes("CBC_J")
colnames(cbc)

# Download Iron Data, check column names
fe <- nhanes("FETIB_J")
colnames(fe)

# Download Arsenic Data, check column names
as <- nhanes("UTAS_J")
colnames(as)

# Download Cadmium and Lead Data, check column names
cd_pb <- nhanes("PBCD_J")
colnames(cd_pb)

# Download Cotinine Data, check column names
cotinine <- nhanes("COT_J")
colnames(cotinine)

# What variable is the same between the datasets?
```

# Subset datasets to keep only variables that we will use in the class

```{r select}

# select() is a useful tidyverse function that keeps columns in a dataframe
# We only want to keep the necessary variables to avoid creating a huge dataset

############ Demographics dataset ############
dim(demog)
#9254   46
demog_subset <- demog %>%
  select(SEQN,     #participant identifier
         RIAGENDR, #sex
         RIDAGEYR, #age in years
         RIDRETH1, #race in number code
         INDFMPIR, #poverty-income ratio
         DMDEDUC3, #education categories for age 6-19
         DMDEDUC2, #education categories for age 20+
         #these two are for later (survey weighting)
         SDMVSTRA, #Strata: based on census region, metropolitan area, or population demographics
         SDMVPSU   #Primary Sampling Unit: 30 per nhanes cycle - mostly single counties, selected from strata
         ) %>%
  rename(RIASEX = RIAGENDR) #rename RIAGENDR to be RIASEX because NHANES asked about sex, not gender
dim(demog)
#9254    9 - number of rows does not change

############ Complete blood count dataset ############
dim(cbc)
#8366   22
cbc_subset <- cbc %>%
  select(SEQN,
         LBXRBCSI,
         LBXWBCSI,
         LBDLYMNO,
         LBDNENO)
dim(cbc)
#8366    2

############ Iron dataset ############
dim(fe)
#6401    9
fe_subset <- fe %>%
  select(SEQN,
         LBXIRN)
dim(fe)
#6401    2

############ Arsenic dataset ############
dim(as)
#2979    4
as_subset <- as %>%
  select(SEQN,
         URXUAS)
dim(as_subset)
#2979    2

############ Cadmium/Lead dataset ############
dim(cd_pb)
#8366   16
cd_pb_subset <- cd_pb %>%
  select(SEQN,
         LBXBCD,
         LBXBPB)
dim(cd_pb_subset)
#8366    3

############ Cotinine dataset ############
dim(cotinine)
#7936    5
cotinine_subset <- cotinine %>%
  select(SEQN,
         LBXCOT)
dim(cotinine)
#7936    2
```


# One dataset at a time, merge all datasets with the demographics dataset

```{r data_merge}

# Merge the demographics and complete blood count datasets
demog_cbc <- left_join(demog_subset, cbc_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc)

# Merge the previous dataset and iron dataset
demog_cbc_fe <- left_join(demog_cbc, fe_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe)

# Merge the previous dataset and arsenic dataset
demog_cbc_fe_as <- left_join(demog_cbc_fe, as_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe_as)

# Merge the previous dataset and cadmium/lead dataset
demog_cbc_fe_as_cdpb <- left_join(demog_cbc_fe_as, cd_pb_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe_as_cdpb)

# Merge the previous dataset and cotinine dataset
# Note, the variable label for SEQN in the cotinine dataset from NHANES has a different capitalization pattern, so R recognizes this as a different variable from the SEQN variable in all of the other datasets. Let's strip the label from that variable and then do the merge.
demog_cbc_fe_as_cdpb <- remove_label(demog_cbc_fe_as_cdpb, SEQN)
cotinine_subset <- remove_label(cotinine_subset, SEQN)
demog_cbc_chems <- left_join(demog_cbc_fe_as_cdpb, cotinine_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_chems)

# This is our merged dataset!
```


# Convert race/ethnicity and sex variables from numeric to factors

```{r convert_variable_types}

# Check the initial counts of race/ethnicity and sex
table(demog_cbc_chems$RIDRETH1)
table(demog_cbc_chems$RIASEX)

# Check the initial data type of the variables
str(demog_cbc_chems$RIDRETH1)
str(demog_cbc_chems$RIASEX)

# Update race/ethnicity and sex from numbers to characters
nhanes_factor <- demog_cbc_chems %>%
  mutate(race_eth = case_when(RIDRETH1 == 1 ~ "Mexican American",
                              RIDRETH1 == 2 ~ "Other Hispanic",
                              RIDRETH1 == 3 ~ "Non-Hispanic White",
                              RIDRETH1 == 4 ~ "Non-Hispanic Black",
                              RIDRETH1 == 5 ~ "Other Race"),
         sex = case_when(RIASEX == 1 ~ "Male",
                         RIASEX == 2 ~ "Female")) %>%
  mutate(sex = factor(sex),               #change the character variables into factor variables
         race_eth = factor(race_eth)) %>%
  mutate(sex = relevel(sex,               #set the reference levels
                       ref = "Male"),
         race_eth = relevel(race_eth,
                            ref = "Non-Hispanic White")) 
# Check the counts of the new variables
table(nhanes_factor$race_eth)
table(nhanes_factor$sex)

# Do the counts still match?

# Check the data type of the new variables
str(nhanes_factor$race_eth)
str(nhanes_factor$sex)
```


# Clean and convert the education variables from numeric to factors
* **https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DEMO_J.htm**

```{r set_factor_levels}

# DMDEDUC3 - participants 6 - 19 years old
# DMDEDUC2 - participants 20+ years old

# Check original counts of DMDEDUC3
table(nhanes_factor$RIDAGEYR < 20, useNA = "always") #3685 participants under 20
table(nhanes_factor$DMDEDUC3, useNA = "always") #values 0-15 and 66
table(nhanes_factor$DMDEDUC3 >= 0, useNA = "always") #2306 participants with education reported

# Check original counts of DMDEDUC2
table(nhanes_factor$RIDAGEYR >= 20, useNA = "always") #5569 participants 20+ years
table(nhanes_factor$DMDEDUC2, useNA = "always") #values 1-5, 7, 9
table(nhanes_factor$DMDEDUC2 >= 0, useNA = "always") #5569 participants with data (13 with 7 or 9)

# Update the numeric codes to categories
nhanes_educ2_3 <- nhanes_factor %>%
  mutate(education_youth = case_when(DMDEDUC3 >= 0 & DMDEDUC3 <= 5 ~ "Less than 5th grade",
                                DMDEDUC3 >= 6 & DMDEDUC3 <= 12 | DMDEDUC3 == 66 ~ "Less than high school",
                                DMDEDUC3 == 13 | DMDEDUC3 == 14 ~ "High school or GED",
                                DMDEDUC3 == 15 ~ "More than high school")) %>%
  mutate(education_adult = case_when(DMDEDUC2 == 1 | DMDEDUC2 == 2 ~ "Less than high school",
                                 DMDEDUC2 == 3 ~ "High school or GED",
                                 DMDEDUC2 == 4 | DMDEDUC2 == 5 ~ "More than high school",
                                 DMDEDUC2 == 7 | DMDEDUC2 == 9 ~ "Unknown")) %>%
  mutate(education_adult = na_if(education_adult, "Unknown")) #Make all the "Unknown" values missing

# Check the counts
table(nhanes_educ2_3$education_youth, useNA = "always")
table(nhanes_educ2_3$education_adult, useNA = "always")


# Merge the two education variables
nhanes_educ <- nhanes_educ2_3 %>%
  mutate(education = coalesce(education_youth,
                              education_adult))

# View the last three columns in the dataset
head(nhanes_educ)

# Set the factor levels and reference for education, reorder the columns
nhanes_educ <- nhanes_educ %>%
  mutate(education = relevel(factor(education,
                                    levels = c("Less than 5th grade",
                                               "Less than high school",
                                               "High school or GED",
                                               "More than high school")),
                             ref = "Less than high school"))

# Check the counts
table(nhanes_educ$education, useNA = "always")
# Note: setting a reference automatically makes the reference the first level
```


# Create a new continuous variable from other continuous variables

```{r create_continuous_variables}

# Create a neutrophil:lymphocyte ratio variable (NLR)
nhanes_nlr <- nhanes_educ %>%
  mutate(nlr = LBDNENO / LBDLYMNO) 


# View the new variable to look at the changes
nhanes_nlr %>%
  select(SEQN,
         LBDNENO,
         LBDLYMNO,
         nlr) %>%
  head()
```


# Create a categorical variable from a continuous variable

```{r create_categorical_variables}

# Create a categorical variable of iron status based on serum concentrations
nhanes_iron_categ <- nhanes_nlr %>%
  mutate(iron_status = case_when(LBXIRN < 60 ~ "Deficient",
                                 LBXIRN > 170 ~ "Excessive",
                                 LBXIRN >= 60 & LBXIRN <= 170 ~ "Normal")) %>%
  mutate(iron_status = relevel(factor(iron_status),
                               ref = "Deficient"))

# Check the structure of the dataset to see that iron_status is after LBXIRN and it is a factor with reference = "Deficient"
str(nhanes_iron_categ)
```


# Use the cut function to slice a numeric variable into groups

```{r cut_continuous_variables}

# Make groups of equal range
nhanes_dataset <- nhanes_iron_categ %>%
  mutate(cut_groups = cut_interval(RIDAGEYR, n = 5)) 
table(nhanes_dataset$cut_groups)


# Other options:

# Five groups with approximately equal numbers of participants (cut_number())
nhanes_age <- nhanes_iron_categ %>%
  mutate(cut_groups = cut_number(RIDAGEYR, n = 5))
table(nhanes_age$cut_groups)
```

# Check your understanding!

Serum cotinine is a continuous variable in our dataset. Let's use it to create a two-level categorical variable for nonsmoking versus smoking. Nonsmoking status can be defined as serum cotinine values less than or equal to 10 ng/mL (Pirkle et al., 1996). 
* Make a new R code chunk
* Make a new factor variable for smoking status
* The variable should have the level "smoking" for cotinine greater than 10
* The variable should have the level "nonsmoking" for cotinine less than or equal to 10
* How many observations are in the nonsmoking category? In the smoking category?



# Save dataset for the future

```{r save_dataset}

# Reorder the variable names for next time
nhanes_col <-  c("SEQN", "RIASEX", "sex", "RIDAGEYR", "cut_groups", "RIDRETH1", "race_eth",  "INDFMPIR", "DMDEDUC3", "DMDEDUC2", "education_youth", "education_adult", "education",       "SDMVSTRA", "SDMVPSU", "LBXRBCSI", "LBXWBCSI", "LBDLYMNO", "LBDNENO", "nlr", "LBXIRN", "iron_status", "URXUAS", "LBXBCD", "LBXBPB", "LBXCOT")
nhanes_dataset <- relocate(nhanes_dataset, all_of(nhanes_col))
head(nhanes_dataset)

# Save dataset as a csv file
write_csv(nhanes_dataset, file = here("nhanes_dataset.csv"))
# Save dataset as an R object
save(nhanes_dataset, file = here("nhanes_dataset.rda"))

# Import dataset back into R from the csv or R object
# Only R object keeps the preset factor variables saved
nhanes_dataset_csv <- read_csv(here(("nhanes_dataset.csv")))
load(here(("nhanes_dataset.rda")))
```


# Remember to save your R script!

# To exit R

```{r exit, eval=F}
# q()
## if you close R, you will be asked to save your workspace image
```