---
title: "EPID 674: Data Management in R"
subtitle: "Homework 2"
author: "List your name here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

```{r load_libraries}
library(nhanesA)
library(tidyverse)
library(sjlabelled)
library(here)
```


# 1. Load data from 2013-2014 NHANES cycle and check the dimensions

```{r load_data}

# Insert code for loading data and checking dimensions here

```
Interpret the output and provide answer to question 1 here.

# 2. Select columns of interest, join the datasets, and filter the participants.

## Select useful variables to keep

```{r select_variables}

# Insert code for selecting the variables of interest to keep

```


## Join the selected datasets together

```{r join_datasets}

# Insert code for joining the datasets

```

##  Filter the joined dataset to participants of interest

```{r filter_dataset}

# Insert code for filtering the participants of interest to keep

```

Interpret the output and provide answer to question 2 here.



# 3. Create categorical variables from continuous variables

## 3a. Create a sex category variable from existing numeric variable

```{r sex_factor_variable}

# Insert code for creating sex category variable here.

```
Interpret the output and provide answer to question 3a here.


## 3b. Create binary mild cognitive impairment variable based on continuous cognition values

```{r mild_cognitive_impairment}

# Insert code for creating cognitive impairment category variable here.

```
Interpret the output and provide answer to question 3b here.

## 3c. Make age category variable

```{r age_groups}

# Insert code for creating age group variable here.

```
Interpret the output and provide answer to question 3c here.


# 4. Save dataset

```{r save_dataset}

# Insert code for saving, loading, and checking the class of dataset here.

```
Interpret the output and provide answer to question 4 here.
