---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Exploring Data with R"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
   html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This is how you set options for all chunks of code

sessionInfo() # check what packages are already loaded by default
```


## Chapter 2, Exploring Data with R

# Identify base packages and install new packages
```{r install_packages, eval=FALSE}

# Install packages. Do this only once.
# Note, we already installed packages to this workspace. If working on your personal computer, will need to run this code
install.packages("sas7bdat")
install.packages("epiDisplay")
install.packages("gmodels")
install.packages("tidyverse")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```


```{r load_packages}

# Load packages. Load relevant packages every time you start a new R session
library(sas7bdat)
library(epiDisplay)
library(gmodels)
library(tidyverse)

sessionInfo() # check what versions of packages you just loaded
```

# Specify file directories
```{r directories}
getwd() # Orient yourself to the default file path format on your computer


directory <- "/cloud/project" #Class option when coding on RStudio Cloud

# Three alternate options to customize and specify source data directory if programming on a personal computer (only need to do the one that works for you)
# directory<-"M:/EPID674/Data" # Alternate option 1: location of the parent dataset in afs space of school computer
# directory <- "/Volumes/GoogleDrive/My Drive/Teaching/EPID674/2020_fall/EPID674_Week2_Class" # Alternate option 2: location of data on my Macbook
# directory<-"C:/Users/bakulski/Documents/EPID674/Data" #Alternate option 3: location of data on a personal PC

date <- format(Sys.Date(), "%Y%m%d")
```


# Load data from SAS, export data for R or excel
```{r load}
# Check file path
file.path(directory, "nhanes3.sas7bdat")

# Load data from SAS
nhanes <- read.sas7bdat(file.path(directory, "nhanes3.sas7bdat"))
colnames(nhanes)
```


```{r save_dataset}
# Save data as an R object
save(nhanes, file = file.path(directory, "nhanes3.rda"))
load(file.path(directory, "nhanes3.rda")) # Try reading in the data that you just saved

# Exporting data as a txt/csv file
write.table(nhanes, file = file.path(directory, "nhanes3.txt"))
write.csv(nhanes, file = file.path(directory, "nhanes3.csv"))

# Try reading in the .csv that you just made
nhanes.csv <- read.csv(file.path(directory, "nhanes3.csv"))
names(nhanes.csv)
```

# Clean up your workspace
```{r workspace}
## list all objects and remove the ones we do not want
ls()
rm(x, y, z)
# rm(list=ls())  #removes all objects
ls()
```

# Explore the NHANES dataset
```{r dataset_explore}
# Explore the data set
class(nhanes) # What type of object is it?
dim(nhanes) # What are the dimensions?
names(nhanes) # What are the names?
colnames(nhanes) # Whate are the column names?
rownames(nhanes)[1:20] # What are the row names?


head(nhanes) # What do the first 6 rows look like?
nhanes[1:10, ] # What do the first 10 rows look like?
```

# Explore individual variables in the NHANES dataset
```{r explore_variables}
# Explore the variables
length(nhanes$age)
nhanes$age[1:10]

length(nhanes$educ)
nhanes$educ[1:10]
```

# Data description functions, numeric variables
```{r descriptives_numeric_variable}
mean(x = nhanes$age, trim = 0, na.rm = FALSE)
mean(na.rm = FALSE, x = nhanes$age, trim = 0)
mean(nhanes$age, 0, F)
mean(nhanes$age)

median(nhanes$age)
quantile(nhanes$age)
quantile(nhanes$age, c(0.1, 0.9))
sd(nhanes$age)
IQR(nhanes$age)

summary(nhanes$age)
summary(nhanes)
```

# Create a factor variable (convert from a numeric variable)
```{r factor_variable}
class(nhanes$sex)
table(nhanes$sex)
sex1 <- factor(nhanes$sex, levels = c(1, 2), labels = c("male", "female")) ## creating sex1 from sex
class(sex1)
is.factor(nhanes$sex)
is.factor(sex1)
levels(sex1)
nhanes$sex[1:10]
sex1[1:10]
table(nhanes$sex)
table(sex1)
nhanes <- cbind(nhanes, sex1) # if the new variable passes QC, you can bind it to the parent dataset

### opposite way is as.numeric()
sex2 <- as.numeric(sex1)
sex2[1:10]
table(sex2)
```

# Calculate descriptive statistics on a categorical variable
```{r categorical_variable_descriptives}
table(nhanes$race)
tab1(nhanes$sex1)
table(nhanes$sex1, nhanes$race)
CrossTable(nhanes$sex1, nhanes$race)
```

# Do Exercise 2A

# Take a random sample of the dataset
```{r random_sample}
## sample() to randomly sample
samp <- sample(nhanes$seqn, 100)
samp[1:20]
samp1 <- sample(nhanes$seqn, 100, replace = T)
samp1[1:20]
```

# Create a subset of the dataset
```{r subset}
# subset(): You can create a subset from a data frame using the subset() function.
# For example, create a dataset for race==1 (White).
table(nhanes$race)
nhanes.white <- subset(nhanes, nhanes$race == 1)
dim(nhanes.white)

# To select only the 'age', 'bmi', and 'sex' variables among Black people
nhanes.black <- subset(nhanes, nhanes$race == 2, select = c(age, bmi, sex))
dim(nhanes.black)

# Use indexing []
nhanes.young <- nhanes[nhanes$age < 30, ]
dim(nhanes.young)
```

# Calculate descriptives on a subset of the dataset
```{r descriptive_statistics_subset} 
summary(nhanes[, 1:5])
summary(nhanes[, c(1, 6, 11, 16, 21)])
summary(nhanes[, c("age", "sex", "race", "bpb", "sbp")])
summary(nhanes[, -c(1:30)])

# Index within objects with [], create subsets, and you can run any R commands for subsets.
mean(nhanes.white$age)
sd(nhanes.black$age)

table(nhanes$race, nhanes$sex)
femalewhite <- nhanes[nhanes$sex == 2 & nhanes$race == 1, ]
nrow(femalewhite)
summary(femalewhite$age)
```

# Use the ifelse() function to create a categorical variable 
```{r ifelse}
# ifelse
length(nhanes$bmi[nhanes$bmi >= 30])
nhanes$obesity <- ifelse(nhanes$bmi >= 30, 1, 0)
tab1(nhanes$obesity)

nrow(nhanes[nhanes$htn == 1 & nhanes$bmi >= 30, ])
nhanes$htnob <- ifelse(nhanes$htn == 1 & nhanes$bmi >= 30, 1, 0)
tab1(nhanes$htnob)
```

# Create a categorical variable from a numeric variable
# Use the cut function to slice a numeric variable into groups
```{r cut}
## create a categorical variable
AGE5a <- cut(nhanes$age, 5) # Five equally spaced intervals (14 yrs)
summary(AGE5a)
AGE5b <- cut(nhanes$age, quantile(nhanes$age, c(0, .2, .4, .6, .8, 1)), include.lowest = T) # Five groups with ~equal number of participants (quintiles)
summary(AGE5b)
AGE5c <- cut(nhanes$age, breaks = c(19, 40, 50, 60, 70, 90)) # Five groups based on pre-specified cut points
summary(AGE5c)
```

# Use the tidyverse to categorize a numeric variable
```{r mutate}
# Tidyverse: another way
nhanes <- nhanes %>% # This is the pipe symbol
  mutate(AGE5d = case_when(
    age < 30 ~ 1,
    age >= 30 & age < 40 ~ 2,
    age >= 40 & age < 50 ~ 3,
    age >= 50 & age < 60 ~ 4,
    age >= 60 ~ 5,
  ))
summary(nhanes$AGE5d)
```

# Categorize a numeric variable the long way
```{r longform_categorization}
# Long form: another way
agecat <- nhanes$age
agecat[nhanes$age < 30] <- "Twenties"
agecat[nhanes$age >= 30 & nhanes$age < 40] <- "Thirties"
agecat[nhanes$age >= 40 & nhanes$age < 50] <- "Fourties"
agecat[nhanes$age >= 50 & nhanes$age < 60] <- "Fifties"
agecat[nhanes$age >= 60 & nhanes$age < 70] <- "Sixties"
agecat[nhanes$age >= 70 & nhanes$age < 80] <- "Seventies"
agecat[nhanes$age >= 80] <- "Eighties"
table(agecat)
```

# Convert categorical variables to numeric
```{r convert_numeric}
age5c <- unclass(AGE5c) # Convert back into numeric
summary(age5c)
class(AGE5c)
class(age5c)

# Bind the variables that are useful to the nhanes dataframe for use later
nhanes <- cbind(nhanes, AGE5b, AGE5c, age5c)
```





# Do Exercise 2B

# Remember to save your R script!
```{r save_modified_dataset}
# Optional, save the nhanes object you modified
save(nhanes, file = file.path(directory, "nhanes3.rda"))
# Or you can save the workspace (includes all objects) with the name of your file
# save.image(file="epid674.RData")
```


# To exit R 
```{r exit, eval=F}
# q()
## if you close R, you will be asked to save your workspace image
```


