---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Exploring Data with R"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
   html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # This is how you set options for all chunks of code

sessionInfo() # check what packages are already loaded by default
```

## Chapter 2, Exploring Data with R

# Identify base packages and install new packages

```{r install_packages, eval=FALSE}

# Install packages. Do this only once.
# Note, we already installed packages to this workspace. If working on your personal computer, will need to run this code
install.packages("tidyverse")
install.packages("here")
install.packages("nhanesA")
install.packages("sjlabelled")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```

```{r load_packages}

# Load packages. Load relevant packages every time you start a new R session
library(tidyverse)
library(here)
library(nhanesA)
library(sjlabelled)
```

# Specify file directories

```{r directories}
here() # Orient yourself to the default file path format on your computer

#Expect to be when coding on RStudio Cloud
#"/cloud/project" 
```

# Data cleaning plan:
1) Download demographics, complete blood counts, and chemical data
2) Check datasets
3) Keep only the useful variables
4) Merge datasets
5) Fix data types and set reference levels
6) Create categorical variable
7) Create age groups
8) Save cleaned dataset


# Import datasets from 2017-2018 NHANES
* Datasets can be found at: **https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017**
* Data are saved in multiple datasets that we will need to combine
* NHANES variables have labels as well as variable names - those labels need to be removed for merging
```{r load}

# Download NHANES demographics data and remove the labels
demog <- nhanes("DEMO_J")
demog_unlab <- remove_label(demog)
# Check the column names (variable names)
colnames(demog_unlab)

# Check the Demographics Data website for information about these variables:
# https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Demographics&CycleBeginYear=2017
#   What does RIAGENDR mean?

# Download Complete Blood Cell Laboratory Data, remove the labels, check the column names
cbc <- nhanes("CBC_J")
cbc_unlab <- remove_label(cbc)
colnames(cbc_unlab)

# Download Iron Data, remove labels, check column names
fe <- nhanes("FETIB_J")
fe_unlab <- remove_label(fe)
colnames(fe_unlab)

# Download Arsenic Data, remove labels, check column names
as <- nhanes("UTAS_J")
as_unlab <- remove_label(as)
colnames(as_unlab)

# Download Cotinine Data, remove labels, check column names
cotinine <- nhanes("COT_J")
cotinine_unlab <- remove_label(cotinine)
colnames(cotinine_unlab)

# Download Cadmium and Lead Data, remove labels, check column names
cd_pb <- nhanes("PBCD_J")
cd_pb_unlab <- remove_label(cd_pb)
colnames(cd_pb_unlab)

# What variable is the same between the datasets?
```


# Explore the NHANES dataset

```{r dataset_explore}

# Explore the data set
str(demog_unlab) # Get detailed information about each variable and the object overall
dim(demog_unlab) # What are the dimensions?
colnames(demog_unlab) # Whate are the column names?
rownames(demog_unlab)[1:20] # What are the row names?


head(demog_unlab) # What do the first 6 rows look like?
demog_unlab[1:10, ] # What do the first 10 rows look like?
```


# Explore individual variables in the NHANES dataset

```{r explore_variables}
# Explore the variables
length(demog_unlab$RIDAGEYR)
demog_unlab$RIDAGEYR[1:10]

# Are there any duplicates?

# Look at the whole dataset
View(demog_unlab)
```

# Subset datasets to keep only useful variables

```{r select()}

# select() is a useful tidyverse function that keeps columns in a dataframe
# We only want to keep the necessary variables to avoid creating a huge dataset

############ Demographics dataset ############
dim(demog_unlab)
#9254   46
demog_subset <- demog_unlab %>%
  select(SEQN,     #participant identifier
         RIAGENDR, #sex
         RIDAGEYR, #age in years
         RIDRETH1, #race in number code
         INDFMPIR, #poverty-income ratio
         #these two are for later (survey weighting)
         SDMVSTRA, #Strata: based on census region, metropolitan area, or population demographics
         SDMVPSU   #Primary Sampling Unit: 30 per nhanes cycle - mostly single counties, selected from strata
         ) %>%
  rename(RIASEX = RIAGENDR) #rename RIAGENDR to be RIASEX because NHANES asked about sex, not gender
dim(demog_subset)
#9254    7 - number of rows does not change

############ Complete blood count dataset ############
dim(cbc_unlab)
#8366   22
cbc_subset <- cbc_unlab %>%
  select(SEQN,
         LBXRBCSI,
         LBXWBCSI,
         LBDLYMNO,
         LBDNENO)
dim(cbc_subset)
#8366    2

############ Iron dataset ############
dim(fe_unlab)
#6401    9
fe_subset <- fe_unlab %>%
  select(SEQN,
         LBXIRN)
dim(fe_subset)
#6401    2

############ Arsenic dataset ############
dim(as_unlab)
#2979    4
as_subset <- as_unlab %>%
  select(SEQN,
         URXUAS)
dim(as_subset)
#2979    2

############ Cotinine dataset ############
dim(cotinine_unlab)
#7936    5
cotinine_subset <- cotinine_unlab %>%
  select(SEQN,
         LBXCOT)
dim(cotinine_subset)
#7936    2

############ Cadmium/Lead dataset ############
dim(cd_pb_unlab)
#8366   16
cd_pb_subset <- cd_pb_unlab %>%
  select(SEQN,
         LBXBCD,
         LBXBPB)
dim(cd_pb_subset)
#8366    3
```


# Merge six datasets into one

```{r data_merge}

# Merge the demographics and complete blood count datasets
demog_cbc <- left_join(demog_subset, cbc_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc)

# Merge the previous dataset and iron dataset
demog_cbc_fe <- left_join(demog_cbc, fe_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe)

# Merge the previous dataset and arsenic dataset
demog_cbc_fe_as <- left_join(demog_cbc_fe, as_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe_as)

# Merge the previous dataset and cotinine dataset
demog_cbc_fe_as_cot <- left_join(demog_cbc_fe_as, cotinine_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_fe_as_cot)

# Merge the previous dataset and cadmium/lead dataset
demog_cbc_chems <- left_join(demog_cbc_fe_as_cot, cd_pb_subset, by = "SEQN")
# Check the merge - How many participants and variables do you expect?
str(demog_cbc_chems)
```


# Convert variables from numeric to factors

```{r}

# Check the initial counts of race/ethnicity and sex
table(demog_cbc_chems$RIDRETH1)
table(demog_cbc_chems$RIASEX)

# Check the initial data type of the variables
str(demog_cbc_chems$RIDRETH1)
str(demog_cbc_chems$RIASEX)

# Update race/ethnicity and sex from numbers to characters
nhanes_factor <- demog_cbc_chems %>%
  mutate(race_eth = case_when(RIDRETH1 == 1 ~ "Mexican American",
                              RIDRETH1 == 2 ~ "Other Hispanic",
                              RIDRETH1 == 3 ~ "Non-Hispanic White",
                              RIDRETH1 == 4 ~ "Non-Hispanic Black",
                              RIDRETH1 == 5 ~ "Other Race"),
         sex = case_when(RIASEX == 1 ~ "Male",
                         RIASEX == 2 ~ "Female")) %>%
  mutate(sex = factor(sex),               #change the character variables into factor variables
         race_eth = factor(race_eth)) %>%
  mutate(sex = relevel(sex,               #set the reference levels
                       ref = "Male"),
         race_eth = relevel(race_eth,
                            ref = "Non-Hispanic White")) %>%
  relocate(race_eth, .after = RIDAGEYR) %>% #reorder the columns to keep the demographics together (mutate() automatically adds the new variable to the end of the dataset)
  relocate(sex, .after = race_eth)
  

# Check the counts of the new variables
table(nhanes_factor$race_eth)
table(nhanes_factor$sex)

# Do the counts still match?

# Check the data type of the new variables
str(nhanes_factor$race_eth)
str(nhanes_factor$sex)
```


# Create a new variable from continuous

```{r}

# Create a neutrophil:lymphocyte ratio variable (NLR)
nhanes_nlr <- nhanes_factor %>%
  mutate(nlr = LBDNENO / LBDLYMNO)


# View the new variable to look at the changes
nhanes_nlr %>%
  dplyr::select(SEQN,
                LBDNENO,
                LBDLYMNO,
                nlr) %>%
  View()
```


# Create a categorical variable from a continuous

```{r categorical}

# Create a categorical variable of iron status based on serum concentrations
nhanes_iron_categ <- nhanes_nlr %>%
  mutate(iron_status = case_when(LBXIRN < 60 ~ "Deficient",
                                 LBXIRN > 170 ~ "Excessive",
                                 LBXIRN >= 60 & LBXIRN <= 170 ~ "Normal")) %>%
  mutate(iron_status = relevel(factor(iron_status),
                               ref = "Deficient")) %>%
  relocate(iron_status, .after = LBXIRN)

# Check the structure of the dataset to see that iron_status is after LBXIRN and it is a factor with reference = "Deficient"
str(nhanes_iron_categ)
```


# Use the cut function to slice a numeric variable into groups
**Three different methods

```{r cut}

# Make groups of equal range
nhanes_age <- nhanes_iron_categ %>%
  mutate(cut_groups = cut_interval(RIDAGEYR, n = 5))
table(nhanes_age$cut_groups)


# Other options:

# Five groups with specified bin widths (16)
nhanes_dataset <- nhanes_iron_categ %>%
  mutate(cut_groups = cut_interval(RIDAGEYR, length = 16))
table(nhanes_age$cut_groups)

# Five groups with approximately equal numbers of participants (cut_number())
nhanes_dataset <- nhanes_iron_categ %>%
  mutate(cut_groups = cut_number(RIDAGEYR, n = 5))
table(nhanes_dataset$cut_groups)
```


# Save dataset for the future

```{r save_dataset}

# Save dataset as a csv file
write.csv(nhanes_age, file = here("nhanes_dataset.csv"),
          row.names=FALSE)
# Save dataset as an R object
save(nhanes_age, file = here("nhanes_dataset.rda"))

# Import dataset back into R from the csv or R object
nhanes_dataset <- read.csv(here(("nhanes_dataset.csv")))
load(here(("nhanes_dataset.rda")))
```


# Remember to save your R script!

# To exit R

```{r exit, eval=F}
# q()
## if you close R, you will be asked to save your workspace image
```